A Computational Account of Default-Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-10-02 paper/paper.tex
wip

Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s_t, a_t)$, a cumulative version which
takes into account the future. A standard for modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  R(s_t, a_t) = r(s_t, s_t) + \ldots + \gamma^Tr(s_{t+T}, s_{t+T})
  = r(s_t, s_t) + \gamma \sum_{\tau=1}^{T}\gamma^{\tau-1}r(s_{t+\tau},a_{t+\tau}),
  \end{equation}
for some \textit{discount factor} $\gamma \in [0, 1]$, and horizon $T$. The choice $\gamma = 0$ (or $T=0$)  corresponds to a perfectly myopic agent who is solely concerned about their immediate rewards and have no regards for the future. For there to be any learning at all, a technical requirement is that $\gamma < 1$ \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. The reward importance hence decays with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount $\gamma$ calibrates the delay discounting
behavior of the intelligent agent.

Note that $R(s_t, a_t)$ defined above is a stochastic quantity which depends on the trajectory generated by the policy $\pi$, upto but not including time $t$
\begin{eqnarray*}
  \begin{split}
    s_0 &\sim \mu\\
    a_0 &\sim \pi(.|s_0)\\
    s_1 &\sim \rho(.|s_0,a_0) \text{ and collect rewards }r(s_0, a_0)\\
    a_1 &\sim \pi(.|s_1)\\
    s_2 &\sim \rho(.|s_1,a_1) \text{ and collect rewards }r(s_1, a_1)\\
    \ldots\\
    a_{t-1} &\sim \pi(.|s_{t-1})\\
    s_{t} &\sim \rho(.|s_{t-1},a_{t-1}) \text{ and collect rewards }r(s_{t-1}, a_{t-1})
  \end{split}
\end{eqnarray*}

%% The goal is then to sample from a policy $a_t \sim \pi$ so that the expected
%% value of $R(s_t, a_t)$ is maximized.
Formally, the agent's goal is then to maximize the state-action value for
taking action $a_t$ and state $s_t$ defined by
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s_t,a_t) = \mathbb E [R(s_t, a_t)]
\end{equation}

where the expectation is over all possible trajectories successor
trajectories of state $s_t$.
Given that each $\pi$ correspondence to an exhaustive description of
behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
the set of actions given states in finite time.

By construction, in general, maximizing the expectation in  \eqref{eq:qval} is intractable due to
the enormous number of possible trajectories to consider.
(If the Bellman-based computational ideas have a corresponding
neurobiological implementation, this provides a framework of how
optimal behavior could be computable in the human brain.)
One popular simplification that removes this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.

\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} [Sutton and Barto 1998] then takes the simple form
\begin{equation}
  Q^\pi(s_t, a_t) = \mathbb E_{s_{t+1} \sim \rho(.|s_t,a_t)} [r(s_t,a_t) + \gamma Q^\pi(s_{t+1}, \pi(s_{t+1}))]
  \label{eq:bellman}
\end{equation}
which is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

@@

\subsection{Correspondence between DMN function and RL framework}
\textbf{XXX Put the diagrams and process charts here.}
\begin{itemize}
  \item The policy matrix represent the repertoire of possible actions
  on the world given a current state. It represents the probabilites of
  choosing actions to be executed in a certain situation.
  \item Given a state, the value function represents the estimated utility of
  being in that state, which is subjective to the agent. 
  For state s there is set of actions a.
  It provides a quantitative answer to the question:
  "What state is desirable for a specific agent."
  For all actions being in one of the states s,
  the Q value function provides the subjective utility
  of executing a specific action.
  It provides a quantitative answer to the question:
  "Which is the best action to take in the current situation."
  \item The DMN may subserve
  constant exploration of possible future actions and their
  cumulative outcomes. Implicit computation of future choices 
  provides an explanation for purposeful mind-wandering.
  \item hippocampus -> provides perturbed action-transition-state-reward samples
  as a block of "imagined", "hypothesized", "recalled" experience;
  the small variations in these experience blocks allow searching
  a larger space of parameters and possible experiences.
  In the absence of environmental input and feedback
  (e.g., mind-wandering or sleep) pseudo-experiencing possible
  future scenarios and action outcomes.

>>>>

Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative version which takes into account the future. A standard for
modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  R^\pi(s_0, a_0) = r(s_0, s_0) + \gamma r(s_1, a_1) + \ldots + 
  = \sum_{t=0}^{\infty}\gamma^{t}r(s_{t},a_{t}),
\end{equation}
which measures the cumulative reward of starting off at state $s_0$ taking action $a_0$, then following a policy $\pi$ for choosing actions henceforth,
i.e

\begin{eqnarray*}
  \begin{split}
    s_0, a_0 &\leftarrow s, a\\
    s_1 &\sim \rho(.|s_0,a_0) \text{ and collect rewards }r(s_0, a_0)\\
    a_1 &\sim \pi(.|s_1)\\
    s_2 &\sim \rho(.|s_1,a_1) \text{ and collect rewards }r(s_1, a_1)\\
    \ldots\\
    a_{t-1} &\sim \pi(.|s_{t-1})\\
    s_{t} &\sim \rho(.|s_{t-1},a_{t-1}) \text{ and collect rewards }r(s_{t-1}, a_{t-1})\\
    \ldots
  \end{split}
\end{eqnarray*}

In \eqref{eq:cumr}, the constant $\gamma \in [0, 1]$ is the \textit{discount factor}. The choice $\gamma = 0$   corresponds to a perfectly myopic agent who is solely concerned about their immediate rewards and have no regards for the future. For there to be any learning at all, a technical requirement is that $\gamma < 1$.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus the agent is positively biased towards considering rewards which are will
come soon. The reward importance hence decays with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount factor $\gamma$ can thus be seen as calibrating
the delay discounting behavior of the intelligent agent.

Note that $R^\pi(s_0, a_0)$ defined above is a stochastic quantity which depends on the trajectory generated by the policy $\pi$ from state $s$ onwards.
The agent's goal is find a polyciy $\pi$ which maximizes the expected cumulative value of state-stat-pairs $(a_0,s_0)$, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s_0,a_0) = \mathbb E [R(s_0, a_0)]
\end{equation}

where the expectation is over all possible trajectories generate by the
policy $\pi$, starting from state $s_0$ and action $a_0$ taken at this
state.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.

By construction, in general, maximizing the expectation in  \eqref{eq:qval} is intractable due to
the enormous number of possible trajectories the agent might need to
consider.
(If the Bellman-based computational ideas have a corresponding
neurobiological implementation, this provides a framework of how
optimal behavior could be computable in the human brain.)
One popular simplification that removes this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.

\textit{Q-learning} (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The celebrated \textit{Bellman's equation} [Sutton and Barto 1998] then takes the simple form
\begin{equation}
  Q^\pi(s, a) = \mathbb E_{s' \sim \rho(.|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))]
  \label{eq:bellman}
\end{equation}
which is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

@@

\subsection{Correspondence between DMN function and RL framework}
\textbf{XXX Put the diagrams and process charts here.}
\begin{itemize}
  \item The policy matrix represent the repertoire of possible actions
  on the world given a current state. It represents the probabilites of
  choosing actions to be executed in a certain situation.
  \item Given a state, the value function represents the estimated utility of
  being in that state, which is subjective to the agent. 
  For a state $s$, the value function at $s$ namely $V(s)$ define provides a quantitative answer to the question:
  "How desirable is it to be in state $s$ ?"
  For each action $a$ that can be taken from state $s$,
  the Q-value  $Q(s, a)$ defined in \eqref{eq:qval} provides the subjective
  utility of executing a specific action; it answers the question
  "What is the expected utiltiy of taking action $a$ in this situation ?"
  \item The DMN may subserve
  constant exploration of possible future actions and their
  cumulative outcomes. Implicit computation of future choices 
  provides an explanation for purposeful mind-wandering.
  \item hippocampus -> provides perturbed action-transition-state-reward samples
  as a block of "imagined", "hypothesized", "recalled" experience;
  the small variations in these experience blocks allow searching
  a larger space of parameters and possible experiences.
  In the absence of environmental input and feedback
  (e.g., mind-wandering or sleep) pseudo-experiencing possible
  future scenarios and action outcomes.
